Wormtable: A data structure for genome scale data

### 2013-07-08 docs/paper/paper.tex
Put in numbers for DGRP discussion.

Wormtable is designed to make working with genome scale data more 
efficient, and in this section we describe some examples of wormtable
that illustrate the improvements over existing methods. We applied these 
examples to a large, publicly available VCF file produced as part of the
"Drosophila Genetic Reference Panel". This VCF ("freeze2.vcf.gz, available 
here: ftp://ftp.hgsc.bcm.edu/DGRP/freeze2_Feb_2013/) contains variant calls 
from a population consisting 192 inbred Drosophila lines and contains data 
from the whole genome over a total of 6,146,611 rows). This VCF was converted to 
wormtable format using the included \texttt{vcf2wt} conversion program.  
The (uncompressed) VCF file was 15Gb, and the converted wormtable was 12Gb,
which required X hours.

Values are stored in wormtable in a binary format, so that no parsing is
required when reading in rows whereas methods that use the VCF directly 
must parse each row before it can be used, and is by far the most time 
consuming part of processing VCF data. To illustrate this advantage, we
wrote a script to count the number of transitions and transversions
in the dataset, using wormtable and using PyVCF[citation?], a module for
python to parse VCF files. In both of these examples we proceed 
row-by-row, checking the REF and ALT columns for the type of variant and
summing transitions and transversions (ignoring indels). The 
PyVCF method required X hours to complete, while the wormtable 
based script needed only Y seconds. This stark difference is not a 
criticism of the PyVCF library, which provides an efficient parser 
for VCF files; it is rather a measure of the inherent slowness of
parsing text in large volumes.

@@

To illustrate the use of indexes in wormtable, we created an index 
on the REF and ALT columns. This index required X MB of disc space 
Y minutes to build, respectively. The REF+ALT index provides a much more 
efficient means of counting the number of transitions and transversions 
(we can just count the number of rows matching keys in the index 
corresponding the SNPs of each type). A script to calculate these 
counts by this method required x seconds.

>>>>

Wormtable is designed to make working with genome scale data more 
efficient, and in this section we describe some examples of wormtable
that illustrate the improvements over existing methods. We applied these 
examples to a large, publicly available VCF file produced as part of the
Drosophila genetic reference panel~\citep{mr+12}.
This VCF\footnote{\href%
{ftp://ftp.hgsc.bcm.edu/DGRP/freeze2\_Feb\_2013/vcf\_files/freeze2.vcf.gz}
{ftp://ftp.hgsc.bcm.edu/DGRP/freeze2\_Feb\_2013/vcf\_files/freeze2.vcf.gz}}
contains variant calls 
from a population consisting 192 inbred Drosophila lines and contains data 
from the whole genome over a total of 6,146,611 rows and consists
of 15GB of uncompressed text. The VCF was converted to 
wormtable format using the included \texttt{vcf2wt} program
on a standard workstation with a single hard disk. The
conversion required approximately 70 minutes, and the size of the resulting  
wormtable file was 10.1GB (using the smallest type required
to represent the data in each column).

Values are stored in wormtable in a binary format, so that no parsing is
required when reading in rows. To illustrate this advantage, we
wrote a script to count the number of transitions and transversions
in the dataset, using wormtable, standard Unix tools and PyVCF\footnote{%
\href{http://pyvcf.rtfd.org/}{http://pyvcf.rtfd.org/}}, 
a Python VCF parser.
In each examples we proceed 
row-by-row, checking the REF and ALT columns for the type of variant and
counting the transitions and transversions we encounter. 
Using PyVCF this required approximately 130 minutes. The Unix tools 
(\texttt{cut}, \texttt{grep} and \texttt{perl}) method required 
1m17s, and wormtable required 55 seconds. Both the Unix tools 
and wormtable were limited by the speed of the hard disk in the
test machine, and therefore could be much faster using 
modern solid state 
storage. It should be noted, however, that this simple task represents 
something of a best case scenario for the Unix tools pipeline.

@@

To illustrate the use of indexes in wormtable, we created an index 
on the REF and ALT columns which required 3m42s to build and 
consumed 66MB of storage space. Using this index we repeated the 
example of counting transitions and transversions, which required 
less than a second to complete.
